---
title: Text Preprocessing using textclean
author: Wulan Andriyani
date: '2019-01-15'
slug: textclean
categories:
  - R
tags:
  - textclean
  - NLP
  - text preprocessiong
  - feature engineering
  - dplyr
  - Capstone Ml
description: ''
featured: 'textclean.png'
featuredalt: ''
featuredpath: 'date'
linktitle: ''
type: post
---


```{r setup, include=FALSE}
# clean up the environment
rm(list = ls())

# setup chunk options
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  comment = "#>"
)
```

# Take tweet from twitter using rtweet package

Function used for take tweet from twitter is search_tweets(). There are several parameter usually used in that function: <br /> 
- `Topic`    : topic you will find at twitter <br />
- `n`           : how many tweet that you want to take <br />
- `include_rts` : logical. If FALSE tweet taken didn't contain retweet <br />
- `lang`        : spesified language. If you want take tweet in english you can add 
                argument lang = "en" <br />
You can use this code below to try taking tweets from twitter by removing the command 

```{r}
library(rtweet)
```

## [Additional] Subset content text from tweets and save that to csv file
```{r}

tweets <- read.csv("data_input/tweets.csv")
```

# Remove duplicate tweet using dplyr package

Function used for retain only unique/distinct rows from input data frame is distinct() 
```{r}
library(dplyr)

tweets_proces <- tweets %>%
  distinct(text)
```

# Text processing (cleaning text) using textclean package

## Check content text
```{r}
library(textclean)

check_text(tweets_proces$text)
```

## Drop empty rows and NA text rows, then change class to character
```{r}
tweets_proces <- tweets_proces %>% 
  drop_empty_row() %>% 
  drop_NA()

tweets_proces <- as.character(tweets$text)
```

## Replace contraction

Contractions to be replaced with their multi-word forms. Example: <br /> 
- wasn't : was not <br />
- I'll   : I will <br />
- isn't  : is not <br />
- i'd    : i would <br />
- etc
```{r}
tweets_proces <- replace_contraction(tweets_proces)
```

## Remove date

Date to be replaced with character. The default is month, date, and year. Example: <br />
- 11-16-1980 : November sixteenth, one thousand nine hundred eighty <br />
- 1/31/2019  : January thirty first, two thousand nineteen <br />
- etc <br />
But, more recommended to add parameters in this function like replacement if you want to replace or remove date content. Example for argument replacement: <br />
replacement = "" : replace date content with blank character or remove date content
```{r}
tweets_proces <- replace_date(tweets_proces, replacement = "")
```

## Remove email

Replaces email addresses with blank character. Example: <br />
- wulan_andriyani@gmail.com : "" <br />
- wulan123@yahoo.co.id      : "" <br />
- etc
```{r}
tweets_proces <- replace_email(tweets_proces)
```

## Replace emoji

Replaces emojis with word equivalents. Example: <br /> 
- <f0><9f><9a><bc> : baby symbol <br />
- <f0><9f><9a><bd> : toilet <br />
- etc <br />
```{r}
tweets_proces <- replace_emoji(tweets_proces)
```

## Replace emoticon

Replaces emoticons with word equivalents. Example: <br />
- :'( : crying <br />
- :)  : smiley <br />
- :*  : kiss <br />
- etc 
```{r}
tweets_proces <- replace_emoticon(tweets_proces)
```

## Replace grade

Replaces grades with word equivalents. Example: <br /> 
- A  : very excellent <br />
- B+ : almost excellent <br />
- D  : bad <br />
- etc
```{r}
tweets_proces <- replace_grade(tweets_proces)
```

## Remove hashtags

Replaces Twitter hash tags with blank character. Example: <br />
- #Rforbigdata      : "" <br />
- #happynewyear2019 : "" <br />
- etc
```{r}
tweets_proces <- replace_hash(tweets_proces)
```

## Remove HTML

Remove angle braces and replaces HTML symbol markup with equivalent symbols. Example: <br />
- <bold>   : "" <br />
- &lt; &gt : < > <br />
- etc <br />
But, more recommended to add parameters in this function like symbol if you want to replace or remove HTML symbol markup. Example for argument symbol: <br />
symbol = TRUE/FALSE : logical. If FALSE HTML symbol markup will be removed  

```{r}
tweets_proces <- replace_html(tweets_proces, symbol = FALSE)
```

## Remove incomplete sentence

Replaces incomplete sentence end marks with "|". Example: <br />
- ... : | <br />
- .?  : | <br />
- etc <br />
But, more recommended to add parameters in this function like replacement if you don't want to add | as replacement of incomplete sentence. Example for argument replacement: <br />
replacement = "" : replace incomplete sentence with blank character or don't add |
```{r}
tweets_proces <- replace_incomplete(tweets_proces, replacement = "")
```

##  Replace internet slang

Replaces internet slang with longer word equivalents that are more easily analyzed. Example: <br />
- 2nte : tonight <br />
- ASAP : as soon as possible <br />
- TGIF : thank god, it's friday <br />
- etc
```{r}
tweets_proces <- replace_internet_slang(tweets_proces)
```

## Replace number

Numeric to be replaced with words. Example: <br />
- 1.997 : November sixteenth, one thousand nine hundred eighty <br />
- 28    : January thirty first, two thousand nineteen <br />
- 1,888 :
- etc <br />
But, more recommended to add parameters in this function like replacement if you want to replace or remove date content. Example for argument replacement: <br />
replacement = "" : replace date content with blank character or remove date content
```{r}
tweets_proces <- replace_number(tweets_proces, remove = TRUE)
```

## Remove tag

Replaces Twitter style handle tags with blank character. Example: <br />
- @ramnath_vaidya : "" <br />
- @wulan123       : "" <br />
- @hadley         : "" <br />
- etc
```{r}
tweets_proces <- replace_tag(tweets_proces)
```

## Remove url

Replace URLs with blank character. Example: <br />
- http://renkun.me/r/2014/07/26/difference-between-magrittr-and-pipeR.html : "" <br />
- ftp://cran.r-project.org/incoming/"                                      : "" <br />
- etc
```{r}
tweets_proces <- replace_url(tweets_proces, replacement = "")
```

# References

[https://cran.r-project.org/web/packages/textclean/textclean.pdf](https://cran.r-project.org/web/packages/textclean/textclean.pdf) 